Fail Faster

### 2025-03-25 paper/paper.tex
figs

\begin{wrapfigure}{r}{.5\textwidth}
\begin{lstlisting}
module Bq : sig
  type 'a t
  val gen_int : int -> int -> int t
  val return : 'a -> 'a t
  val bind : 'a t -> ('a -> 'b t) -> 'b t

@@

  val fixed_point : ('a t -> 'a t) -> 'a t
  $\dots$
end
\end{lstlisting}
\caption{Some functions from the API of the \bq{} generator DSL.}
\label{fig:bq-api}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.5\textwidth}
\begin{lstlisting}
let tree_of g = fixed_point (fun rg ->
  let%bind n = size in
  weighted_union [
    (1, return E);
    (n,
      let%bind x = g in
      let%bind l = with_size (n / 2) rg in
      let%bind r = with_size (n / 2) rg in
      return (Node (l,x,r)))
  ])
\end{lstlisting}
\caption{A generator using a variety of convenience functions.}\label{fig:tree_of}
\end{wrapfigure}

@@

The basic functionality of the staged monadic generator DSL can be
found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that ignores its \texttt{size} and
\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation
\texttt{k} only gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size random}
and \texttt{k .<a>. size random} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time---they may themselves be the results of calling \texttt{Gen.int}---the arguments 
\texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This is just a 2-argument version of ``The Trick''
from Section~\ref{subsection:msp}.
\ifaftersubmission
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

\begin{figure}
\begin{lstlisting}
module Gen = struct
  type 'a t = int code -> Random.t code -> 'a code

@@

@@

\begin{figure}
\begin{lstlisting}
let int_pair_staged : (int * int) Gen.t =
  Gen.bind (Gen.int .<0>. .<100>.) (fun cx ->
    Gen.bind (Gen.int .<0> cx) (fun cy ->
      Gen.return .<(.~cx,.~cy)>.
    )
  )

@@

\begin{figure}
\begin{lstlisting}
module Gen =
$\dots$
  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) size random : 'a code =
    match weighted_gens with
    | [] -> .< failwith "Error" >.
    | (wc,g) :: gens' ->
      .<
        if .~acc <= .~wc then .~(g size random)
        else
          let acc' = .~acc - .~wc in
          .~(pick .<acc'>. gens' size random)
      >.

@@

\begin{figure}
\begin{lstlisting}
let grades : char Bq.t = Gen.to_bq (
  Gen.bind size (fun n ->
    Gen.weighted_union [
      (n, Gen.return .<'a'>.);
      (.<2>., Gen.return .<'b'>.);
      (.<1>., Gen.return .<'c'>.);
    ]
  )
)
(*
.< fun size random ->
    let sum = size + 2 + 1 + 0 in
    let r = SR.int random 0 sum in
    if r <= size then 'a' else
      let r' = r - size in
      if r' <= 2 then 'b' else
        let r'' = r' - 2 in
        if r'' <= 1 then 'c' else failwith "Error"
>.
*)
\end{lstlisting}
\caption{Use of Staged Weighted Union}
\label{fig:staged-weighted-union-example}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}
module CodeCps = struct
  type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }

@@

\begin{figure}
\begin{lstlisting}
type 'a handle
val recurse : 'a handle -> 'a code Gen.t
val fixed_point : ('a handle -> 'a code Gen.t) -> 'a code Gen.t
\end{lstlisting}
\caption{Staged Recursive Generator Combinator API}
\label{fig:staged-recursive-generator}
\end{figure}

>>>>

\begin{figure}[h]
\begin{lstlisting}
module Bq : sig
  type 'a t
  val gen_int : int -> int -> int t
  val return : 'a -> 'a t
  val bind : 'a t -> ('a -> 'b t) -> 'b t

@@

  val fixed_point : ('a t -> 'a t) -> 'a t
  $\dots$
end
\end{lstlisting}
\caption{Some functions from the API of the \bq{} generator DSL.}
\label{fig:bq-api}
\end{figure}

@@

\begin{wrapfigure}{r}{.49\textwidth}
  \vspace{1.5em}
\begin{lstlisting}[
    xleftmargin=0pt,
    xrightmargin=0pt,
    aboveskip=0pt,
    belowskip=0pt,
    basicstyle=\ttfamily\footnotesize,
    lineskip=0pt
  ]
let tree_of g = fixed_point (fun rg ->
  let%bind n = size in
  weighted_union [
    (1, return E);
    (n,
      let%bind x = g in
      let%bind l = with_size (n / 2) rg in
      let%bind r = with_size (n / 2) rg in
      return (Node (l,x,r)))
  ])
\end{lstlisting}
\caption{A generator using a variety of convenience functions.}\label{fig:tree_of}
\end{wrapfigure}

@@

The basic functionality of the staged monadic generator DSL can be
found in Figure~\ref{fig:gen-staged-basic}.

\begin{figure}[h]
  \begin{lstlisting}
  module Gen = struct
    type 'a t = int code -> Random.t code -> 'a code

@@

The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that ignores its \texttt{size} and
\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation
\texttt{k} only gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size random}
and \texttt{k .<a>. size random} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time---they may themselves be the results of calling \texttt{Gen.int}---the arguments 
\texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This is just a 2-argument version of ``The Trick''
from Section~\ref{subsection:msp}.
\ifaftersubmission
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

\begin{figure}[h]
\begin{lstlisting}
let int_pair_staged : (int * int) Gen.t =
  Gen.bind (Gen.int .<0>. .<100>.) (fun cx ->
    Gen.bind (Gen.int .<0> cx) (fun cy ->
      Gen.return .<(.~cx,.~cy)>.
    )
  )

@@

\begin{figure}[h]
\begin{lstlisting}
module Gen =
$\dots$
  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) size random : 'a code =
    match weighted_gens with
    | [] -> .< failwith "Error" >.
    | (wc,g) :: gens' ->
      .<
        if .~acc <= .~wc then .~(g size random)
        else
          let acc' = .~acc - .~wc in
          .~(pick .<acc'>. gens' size random)
      >.

@@

\begin{figure}[h]
\begin{lstlisting}
let grades : char Bq.t = Gen.to_bq (
  Gen.bind size (fun n ->
    Gen.weighted_union [
      (n, Gen.return .<'a'>.);
      (.<2>., Gen.return .<'b'>.);
      (.<1>., Gen.return .<'c'>.);
    ]
  )
)
(*
.< fun size random ->
    let sum = size + 2 + 1 + 0 in
    let r = SR.int random 0 sum in
    if r <= size then 'a' else
      let r' = r - size in
      if r' <= 2 then 'b' else
        let r'' = r' - 2 in
        if r'' <= 1 then 'c' else failwith "Error"
>.
*)
\end{lstlisting}
\caption{Use of Staged Weighted Union}
\label{fig:staged-weighted-union-example}
\end{figure}

@@

\begin{figure}[h]
  \begin{lstlisting}
  module CodeCps = struct
    type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }

@@

\begin{figure}[h]
\begin{lstlisting}
type 'a handle
val recurse : 'a handle -> 'a code Gen.t
val fixed_point : ('a handle -> 'a code Gen.t) -> 'a code Gen.t
\end{lstlisting}
\caption{Staged Recursive Generator Combinator API}
\label{fig:staged-recursive-generator}
\end{figure}
